Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-10 arxiv/proof_examples.tex
introduction wip;
section 2 wip
section 2 restructure

\subsection{The Random-Effects Model}\label{app_iso_cov}

We provide the proof of Corollary \ref{cor_MTL_loss}.

@@

First, we show that $g(x)$ and $\ex{g(x) \mid X^{(1)}, X^{(2)}}$ are close using standard concentration bounds.
\begin{claim}\label{claim_largedev1}
 In the setting of Corollary \ref{cor_MTL_loss}, we have that with high probability
$$\left| g(x) - \exarg{\epsilon^{(1)}, \epsilon^{(2)} ,\beta_1,\beta_2}{g(x) \mid X_1, X_2 }\right| \le p^{1/2+c} \left(\sigma^2 + \kappa^2+d^2 \right).$$
%for any small constant $c>0$.
 \end{claim}
 \begin{proof}
By Fact \ref{lem_minv} (iii), we have that with high probability,
 \be\label{op_X12}
\|X^{(1)}\|\le \sqrt{(\sqrt{n_1} + \sqrt{p})^2 + n_1 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, \quad \|X^{(2)}\|\le \sqrt{(\sqrt{n_2} + \sqrt{p})^2 + n_2 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p},
 \ee
 and
 \be\label{op_Sig1}
\| \hat\Sigma(x)^{-1}\|\le \frac1{x^2[(\sqrt{n_1} - \sqrt{p})^2 - n_1 \cdot p^{-c_{\varphi}}]+[(\sqrt{n_2} - \sqrt{p})^2 - n_2 \cdot p^{-c_{\varphi}}] }\lesssim \frac{1}{(x^2+1)p},
 \ee
where we used that $1+\tau\le \rho_1=n_1/p\le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p\le \tau^{-1}$ for a small constant $\tau>0$.

@@

\begin{proof}
Corresponding to equation \eqref{EgA}, we define the function
\begin{align*}
h(x):= [(x-1)^2\kappa^2 + (x^2+1)d^2/2] \cdot p^{-1} \tr\left[ (X^{(1)})^\top X^{(1)} \hat\Sigma(x)^{-1}(X^{(2)})^\top X^{(2)}\right].
\end{align*}
Using Fact \ref{lem_minv} (iii), we can obtain that with high probability,
\begin{align}
p(x^2+1)^{-1}&\lesssim \left( x^2 \left[(\sqrt{n_2}-\sqrt{p})^{2}-n_2p^{-c_\varphi}\right]^{-1} +\left[(\sqrt{n_1}-\sqrt{p})^{2}-n_1p^{-c_\varphi}\right]^{-1} \right)^{-1} \nonumber\\
&\preceq (X^{(1)})^\top X^{(1)} \hat\Sigma(x)^{-1}(X^{(2)})^\top X^{(2)} =\left( x^2 [(X^{(2)})^\top X^{(2)}]^{-1} + [(X^{(1)})^\top X^{(1)}]^{-1} \right)^{-1} \nonumber\\
&\preceq \left( x^2 \left[(\sqrt{n_2}+\sqrt{p})^{2}+n_2p^{-c_\varphi}\right]^{-1} +\left[(\sqrt{n_1}+\sqrt{p})^{2}+n_1p^{-c_\varphi}\right]^{-1} \right)^{-1}\lesssim p(x^2+1)^{-1}, \label{op_uplow}
\end{align}
where we also used that $1+\tau\le \rho_1=n_1/p \le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p \le \tau^{-1}$ for a small constant $\tau>0$. Then we get that
\be\label{eq_h1}h(1)= d^2 \cdot p^{-1} \tr\left[ (X^{(1)})^\top X^{(1)} \hat\Sigma(1)^{-1}(X^{(2)})^\top X^{(2)}\right]\lesssim pd^2 .\ee
On the other hand, if $|x-1|\ge \delta$ for a small constant $\delta>0$, then we have
\begin{align}\label{eq_hA}
h(x)\gtrsim (1+x^2)\kappa^2 \cdot p\left( x^2 +1\right)^{-1} =p\kappa^2 ,
\end{align}
where in the first step we used that $|x-1|^2\gtrsim 1+x^2$ for $|x-1|\ge \delta$ and \eqref{op_uplow}.
%and in the second step we used equation \eqref{choiceofpara0} \HZ{I'm not sure which condition we use? put it here}.
Hence using Claim \ref{claim_largedev1} and equations \eqref{eq_h1}-\eqref{eq_hA}, we obtain that with high probability,
\begin{align}
g(x)&\ge  h(x) +\sigma^2(n_1+n_2-p) -p^{-1/2+e_1}\cdot p\left(\sigma^2 +\kappa^2+d^2 \right) \nonumber\\
&\ge h(1) +\sigma^2(n_1+n_2-p) + p^{-1/2+e_1}\cdot p\left(\sigma^2 +\kappa^2+d^2 \right) \ge g(1), \label{gA>g1}
\end{align}
under conditions $\sigma^2 \lesssim  \kappa^2$ and $d^2 \le p^{-c_{\varphi}}{\kappa^2}$. %\eqref{choiceofpara0} \HZ{again, I don't see which condition we use here?}.
This gives that
%\be\label{rough A-1}
$|\hat x-1|\le \delta$ for any small constant $\delta>0$.

@@

On the other hand, we have that with high probability,
\begin{align}
 \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\hat \Sigma^{-2}\right)_{ii}  &\le {p}^{-1}\tr\left[\hat\Sigma^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] \label{eq_bias111}\\ 
 &\le \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\hat\Sigma^{-2}\right)_{ii} .\nonumber
\end{align}
To obtain this inequality, we used  
$$ (\sqrt{n_1}-\sqrt{p})^4 \cdot (1-p^{-c_\varphi}) \preceq \left((X^{(1)})^\top X^{(1)} \right)^2 \preceq (\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) \quad \text{with high probability},$$
by Fact \ref{lem_minv} (iii), and the fact that for the product of two PSD matrices, its trace is always nonnegative.
Using \eqref{lem_cov_derv_eq} with $\Sigma^{(1)}=\Sigma^{(2)}=\Lambda=V=\id_p$ and $w$ being the $i$-th coordinate vector, we can calculate that 
\begin{align}\label{eq_bias112}
 \left(\hat\Sigma^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{a_3+a_4+1}{(a_1+a_2)^2} +\OO(p^{-c_\varphi})
\end{align}
with high probability. Solving equation \eqref{eq_a34extra} with $a_1, a_2$ in \eqref{simplesovlea12} and $\lambda_i\equiv 1$, $1\le i\le p$, we can obtain that
	\begin{align*}
		% a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad & a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}. %\label{simplesovlea34}
	\end{align*}
Inserting them into \eqref{eq_bias112}, we obtain that 
\begin{align*}%\label{eq_bias113}
 \left(\hat\Sigma^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} +\OO(p^{-c_\varphi})
\end{align*}
with high probability. In fact, we need this estimate to hold simultaneously for all $1\le i \le p$ with high probability. For this purpose, we shall use \eqref{apply derivlocal} to get that 
\begin{align*}%\label{eq_bias113}
\left| \left(\hat\Sigma^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{-\frac{\varphi-4}{2\varphi}}
\end{align*}
on a high probability event that does not depend on $i$. Then using Fact \ref{lem_stodomin} (i), we obtain that 
\begin{align*} 
\left| \sum_{i=1}^p \left(\hat\Sigma^{-2}\right)_{ii} - \frac{p}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{1-\frac{\varphi-4}{2\varphi}}
\end{align*}
with high probability, which by Definition \ref{stoch_domination} gives that
\begin{align}\label{eq_bias113}
\left| p^{-1}\sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\le p^{-c_\varphi}
\end{align}
with high probability. Inserting \eqref{eq_bias113} into \eqref{eq_bias111}, we get that
\begin{align*}
		& \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\nonumber \\
		\le &  {p}^{-1}\tr\left[\hat\Sigma^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] - \frac{n_1^2}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} \nonumber\\
		\le &  \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}
\end{align*}
with high probability.

>>>>

\section{Proof of Corollary \ref{cor_MTL_loss}}\label{app_iso_cov}

We follow a similar logic from Theorem \ref{thm_many_tasks}.

@@

First, we show that $g(x)$ and $\ex{g(x) \mid X^{(1)}, X^{(2)}}$ are close using standard concentration bounds.
\begin{claim}\label{claim_largedev1}
 In the setting of Corollary \ref{cor_MTL_loss}, we have that with high probability
$$\left| g(x) - \exarg{\epsilon^{(1)}, \epsilon^{(2)} ,\beta_1,\beta_2}{g(x) \mid X_1, X_2 }\right| \le p^{1/2+c} \left(\sigma^2 + \kappa^2+d^2 \right).$$
%for any small constant $c>0$.
 \end{claim}
 \begin{proof}
By Fact \ref{fact_minv}(ii), we have that with high probability,
 \be\label{op_X12}
\|X^{(1)}\|\le \sqrt{(\sqrt{n_1} + \sqrt{p})^2 + n_1 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p}, \quad \|X^{(2)}\|\le \sqrt{(\sqrt{n_2} + \sqrt{p})^2 + n_2 \cdot p^{-c_{\varphi}}}\lesssim \sqrt{p},
 \ee
 and
 \be\label{op_Sig1}
\| \hat\Sigma(x)^{-1}\|\le \frac1{x^2[(\sqrt{n_1} - \sqrt{p})^2 - n_1 \cdot p^{-c_{\varphi}}]+[(\sqrt{n_2} - \sqrt{p})^2 - n_2 \cdot p^{-c_{\varphi}}] }\lesssim \frac{1}{(x^2+1)p},
 \ee
where we used that $1+\tau\le \rho_1=n_1/p\le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p\le \tau^{-1}$ for a small constant $\tau>0$.

@@

\begin{proof}
Corresponding to equation \eqref{EgA}, we define the function
\begin{align*}
h(x):= [(x-1)^2\kappa^2 + (x^2+1)d^2/2] \cdot p^{-1} \tr\left[ (X^{(1)})^\top X^{(1)} \hat\Sigma(x)^{-1}(X^{(2)})^\top X^{(2)}\right].
\end{align*}
Using Fact \ref{fact_minv}(ii), we can obtain that with high probability,
\begin{align}
p(x^2+1)^{-1}&\lesssim \left( x^2 \left[(\sqrt{n_2}-\sqrt{p})^{2}-n_2p^{-c_\varphi}\right]^{-1} +\left[(\sqrt{n_1}-\sqrt{p})^{2}-n_1p^{-c_\varphi}\right]^{-1} \right)^{-1} \nonumber\\
&\preceq (X^{(1)})^\top X^{(1)} \hat\Sigma(x)^{-1}(X^{(2)})^\top X^{(2)} =\left( x^2 [(X^{(2)})^\top X^{(2)}]^{-1} + [(X^{(1)})^\top X^{(1)}]^{-1} \right)^{-1} \nonumber\\
&\preceq \left( x^2 \left[(\sqrt{n_2}+\sqrt{p})^{2}+n_2p^{-c_\varphi}\right]^{-1} +\left[(\sqrt{n_1}+\sqrt{p})^{2}+n_1p^{-c_\varphi}\right]^{-1} \right)^{-1}\lesssim p(x^2+1)^{-1}, \label{op_uplow}
\end{align}
where we also used that $1+\tau\le \rho_1=n_1/p \le \tau^{-1}$ and $1+\tau\le \rho_2=n_2/p \le \tau^{-1}$ for a small constant $\tau>0$. Then we get that
\be\label{eq_h1}h(1)= d^2 \cdot p^{-1} \tr\left[ (X^{(1)})^\top X^{(1)} \hat\Sigma(1)^{-1}(X^{(2)})^\top X^{(2)}\right]\lesssim pd^2 .\ee
On the other hand, if $|x-1|\ge \delta$ for a small constant $\delta>0$, then we have
\begin{align}\label{eq_hA}
h(x)\gtrsim (1+x^2)\kappa^2 \cdot p\left( x^2 +1\right)^{-1} =p\kappa^2 ,
\end{align}
where in the first step we used that $|x-1|^2\gtrsim 1+x^2$ for $|x-1|\ge \delta$ and \eqref{op_uplow}.
%and in the second step we used equation \eqref{choiceofpara0} \HZ{I'm not sure which condition we use? put it here}.
Hence using Claim \ref{claim_largedev1} and equations \eqref{eq_h1}-\eqref{eq_hA}, we obtain that with high probability,
\begin{align}
g(x)&\ge  h(x) +\sigma^2(n_1+n_2-p) -p^{-1/2+e_1}\cdot p\left(\sigma^2 +\kappa^2+d^2 \right) \nonumber\\
&\ge h(1) +\sigma^2(n_1+n_2-p) + p^{-1/2+e_1}\cdot p\left(\sigma^2 +\kappa^2+d^2 \right) \ge g(1), \label{gA>g1}
\end{align}
under conditions $\sigma^2 \lesssim  \kappa^2$ and $d^2 \le p^{-c_{\varphi}}{\kappa^2}$. %\eqref{choiceofpara0} \HZ{again, I don't see which condition we use here?}.
This gives that
%\be\label{rough A-1}
$|\hat x-1|\le \delta$ for any small constant $\delta>0$.

@@

On the other hand, we have that with high probability,
\begin{align}
 \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\hat \Sigma^{-2}\right)_{ii}  &\le {p}^{-1}\tr\left[\hat\Sigma^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] \label{eq_bias111}\\ 
 &\le \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi})}{p}\sum_{i=1}^p \left(\hat\Sigma^{-2}\right)_{ii} .\nonumber
\end{align}
To obtain this inequality, we used  
$$ (\sqrt{n_1}-\sqrt{p})^4 \cdot (1-p^{-c_\varphi}) \preceq \left((X^{(1)})^\top X^{(1)} \right)^2 \preceq (\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) \quad \text{with high probability},$$
by Fact \ref{fact_minv}(ii), and the fact that for the product of two PSD matrices, its trace is always nonnegative.
Using \eqref{lem_cov_derv_eq} with $\Sigma^{(1)}=\Sigma^{(2)}=\Lambda=V=\id_p$ and $w$ being the $i$-th coordinate vector, we can calculate that 
\begin{align}\label{eq_bias112}
 \left(\hat\Sigma^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{a_3+a_4+1}{(a_1+a_2)^2} +\OO(p^{-c_\varphi})
\end{align}
with high probability. Solving equation \eqref{eq_a34extra} with $a_1, a_2$ in \eqref{simplesovlea12} and $\lambda_i\equiv 1$, $1\le i\le p$, we can obtain that
	\begin{align*}
		% a_1 = \frac{\rho_1(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} ,\quad & a_2 = \frac{\rho_2(\rho_1 + \rho_2 - 1)}{(\rho_1 + \rho_2)^2} , \label{simplesovlea12}\\
		a_3 = \frac{\rho_1}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}, \quad
		&  a_4 = \frac{\rho_2}{(\rho_1 + \rho_2)(\rho_1 + \rho_2 - 1)}. %\label{simplesovlea34}
	\end{align*}
Inserting them into \eqref{eq_bias112}, we obtain that 
\begin{align*}%\label{eq_bias113}
 \left(\hat\Sigma^{-2}\right)_{ii} = \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} +\OO(p^{-c_\varphi})
\end{align*}
with high probability. In fact, we need this estimate to hold simultaneously for all $1\le i \le p$ with high probability. For this purpose, we shall use \eqref{apply derivlocal} to get that 
\begin{align*}%\label{eq_bias113}
\left| \left(\hat\Sigma^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{-\frac{\varphi-4}{2\varphi}}
\end{align*}
on a high probability event that does not depend on $i$. Then using Fact \ref{lem_stodomin} (i), we obtain that 
\begin{align*} 
\left| \sum_{i=1}^p \left(\hat\Sigma^{-2}\right)_{ii} - \frac{p}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\prec p^{1-\frac{\varphi-4}{2\varphi}}
\end{align*}
with high probability, which by Definition \ref{stoch_domination} gives that
\begin{align}\label{eq_bias113}
\left| p^{-1}\sum_{i=1}^p \left(\Sigma(1)^{-2}\right)_{ii} - \frac{1}{(n_1+n_2)^2}\cdot \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\right|\le p^{-c_\varphi}
\end{align}
with high probability. Inserting \eqref{eq_bias113} into \eqref{eq_bias111}, we get that
\begin{align*}
		& \frac{(\sqrt{n_1}-\sqrt{p})^4 \cdot (1- p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}\nonumber \\
		\le &  {p}^{-1}\tr\left[\hat\Sigma^{-2} \left((X^{(1)})^\top X^{(1)} \right)^2\right] - \frac{n_1^2}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3} \nonumber\\
		\le &  \frac{(\sqrt{n_1}+\sqrt{p})^4 \cdot (1+ p^{-c_\varphi}) - n_1^2 \cdot (1+\OO(p^{-c_\varphi}))}{(n_1+n_2)^2}\cdot  \frac{(\rho_1+\rho_2)^3}{(\rho_1+\rho_2-1)^3}
\end{align*}
with high probability.
